\chapter{Introduction}

\section{Current results}

In this section we will explore the current state of the complexity analysis as $r$ and $\ell$ grows. As we fullfill the following table we expose the strategies and how those can be used to enlight the more complex results.

The most trivial result is the recognition of the $(1,0)$-graphs, as in order to recognize it we just need to know if $|E(G)| > 0$. Therefore it's complexity is \O{1}.

\begin{table}[h!]
  \caption{Incomplete complexity analysis of the \RL-partite recognition problem}
  \label{}
  \center
  \begin{tabular}{l|*{6}c}
    \toprule
    \backslashbox{$r$}{$\ell$} & 0 & 1 & 2 & 3 & 4 & \ldots \\
    \midrule
    0 & -  & \? & \? & \? & \? & \ldots \\
    1 & \O{1} & \? & \? & \? & \? & \ldots \\
    2 & \? & \? & \? & \? & \? & \ldots \\
    3 & \? & \? & \? & \? & \? & \ldots \\
    4 & \? & \? & \? & \? & \? & \ldots \\
    $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ \\
  \end{tabular}
\end{table}

\subsection{$m$-bounded results}

Naturally, we wish to find the complexity for those problems of small partition cardinality.

At \cite{konig36}, König showed that it takes \O{m} steps to recognize a bipartite graph. For complete graphs, is enough to check if $|E(G)|=n(n-1)/2$, if it doesn't then we have an answer; If it does, checking vertex by vertex if it's neighborhood contains all other vertex is \O{m}.

For co-bipartite graphs recognition, we need only to verify if it's complement is a bipartite graph. The recognition of a split graph can be done using their vertex degrees \cite{??}, obtaining all vertices degrees is \O{m} therefore the recognition of split graphs is \O{m}
% \begin{table}[h!]
%   \caption{Incomplete complexity analysis of the \RL-partite recognition problem}
%   \label{}
%   \center
%   \begin{tabular}{l|*{6}c}
%     \toprule
%     \backslashbox{$r$}{$\ell$} & 0 & 1 & 2 & 3 & 4 & \ldots \\
%     \midrule
%     0 & -  & \O{m} & \O{m} & \? & \? & \ldots \\
%     1 & \O{1} & \O{m} & \? & \? & \? & \ldots \\
%     2 & \O{m} & \? & \? & \? & \? & \ldots \\
%     3 & \? & \? & \? & \? & \? & \ldots \\
%     4 & \? & \? & \? & \? & \? & \ldots \\
%     $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ \\
%   \end{tabular}
% \end{table}

\subsection{$NP$-Complete results}

A adequate strategy at this moment is to find when the recognition problem gets $NP$-Complete.

At \cite{gareyjohnson} is shown that 3-coloring a graph (i.e. assign a color between three possibles to each vertex such that no neighborhood repeats a color) is $NP$-Complete, it's trivial to see how the 3-coloring of a graph can be reduced to the problem of finding if a graph is a $(3,0)$-graph, therefore the recognition of $(3,0)$-graphs is $NP$-Complete.

It's noticible that the recognition of \RL-graphs is monotonic, and therefore if the recognition of $(3,0)$-graphs are $NP$-Complete then the recognition of any $(r,0)$-graph or $(3,\ell)$-graph is $NP$-Complete for $r > 3$ and $\ell >0$.

We can extrapolate these findings and argument that the recognition of a $(0,3)$-graph is also $NP$-Complete, as it is the same as recognize it's complement as a $(3,0)$-graph, and use the property of monotonicity to state that the recognition of any $(r,3)$-graph or $(0,\ell)$-graph is $NP$-Complete for $r > 0$ and $\ell >3$.

\begin{table}[h!]
  \caption{Incomplete complexity analysis of the \RL-partite recognition problem}
  \label{}
  \center
  \begin{tabular}{l|*{6}c}
    \toprule
    \backslashbox{$r$}{$\ell$} & 0 & 1 & 2 & 3 & 4 & \ldots \\
    \midrule
    0 & -  & \O{m} & \O{m} & \NPc & \NPc & \ldots \\
    1 & \O{1} & \O{m} & \? & \NPc & \NPc & \ldots \\
    2 & \O{m} & \? & \? & \NPc & \NPc & \ldots \\
    3 & \NPc & \NPc & \NPc & \NPc & \NPc & \ldots \\
    4 & \NPc & \NPc & \NPc & \NPc & \NPc & \ldots \\
    $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ \\
  \end{tabular}
\end{table}

\subsection{Frontier results}

Finally, the frontier cases $(1,2)$,$(2,1)$ and $(2,2)$ were subject of studies by Brandstädt\cite{brand-84,brand-96}. He's findings show that:
\begin{itemize}
  \item Recognition of $(1,2)$-graphs are \O{n^4}.
  \item Recognition of $(2,1)$-graphs are \O{n^4}.
  \item Recognition of $(2,2)$-graphs are \O{n^{12}}.
\end{itemize}

\begin{table}[h!]
  \caption{Current complexity analysis of the \RL-partite recognition problem}
  \label{table:current-values}
  \center
  \begin{tabular}{l|*{6}c}
    \toprule
    \backslashbox{$r$}{$\ell$} & 0 & 1 & 2 & 3 & 4 & \ldots \\
    \midrule
    0 & -  & \O{m} & \O{m} & \NPc & \NPc & \ldots \\
    1 & \O{1} & \O{m} & \O{n^4} & \NPc & \NPc & \ldots \\
    2 & \O{m} & \O{n^4} & \O{n^{12}} & \NPc & \NPc & \ldots \\
    3 & \NPc & \NPc & \NPc & \NPc & \NPc & \ldots \\
    4 & \NPc & \NPc & \NPc & \NPc & \NPc & \ldots \\
    $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ \\
  \end{tabular}
\end{table}

\section{On the recognition of $(2,1)$-graphs}

In this section we will describe a algorithm to recognize a $(2,1)$-graph.

Let $G$ be a graph which we wish to know if posses a $(2,1)$-partition. In order for $G$ to accept a $(2,1)$-partition a necessary condition is that no vertex $v \in V(G)$ can spawn a open neighborhood $N$ which does not induces a split graph, or, the vertices not in $N$ should induce a bipartite graph.

This conditions leads us to the following algorithm:

\begin{code}{Is not (2,1)-partitionated}
  def is_Not_21(Graph G) Bool
    for v in V(G) begin
      N = open_neighborhood(v)
      K = V(G) - N
      if (!is_Split(N) and !is_Bipartite(K))
        return true;
    end
    return false;
  end
\end{code}

A `false` return from the algorithm described above, doesn't indicates that the graph is a $(2,1)$-graph, but raises a interesting condition.

Let $A$ be the set of vertices that posses the property where it's open neighborhood($N$) induces a split graph in $G$, and $B$ the set of vertices that the not-neighborhood induces a bipartite graph. If $A \cap B = \emptyset$ then the graph is a $(2,1)$-graph; Otherwise the graph is guarantee to be a $(3,1)$-graph, since there's a vertex which is capable of delimiting 3 disjointed stable sets (Both from the one induced by the not-neighborhood and the on induced by the neighborhood) and a clique (induced by the neighborhood). Then the remaining work shall be to determine if the $(3,1)$-graph is a $(2,1)$-graph.

\begin{code}{Get a (3,1)-partition}
  def get_31(Graph G) 3-1-Partition
    for v in V(G) begin
      N = open_neighborhood(v)
      K = V(G) - N
      if (is_Split(N) and is_Bipartite(K))
        return new 3-1-Partition{
          R1: K.R1,
          R2: K.R2,
          R3: N.R,
          L: N.L
          };
    end
    return empty31();
  end
\end{code}
